---
title: "An end-to-end ChIP-seq workflow: from a peak BED to enriched motifs"
shorttitle: "ChIP-seq workflow"
date: 23 May 2026
author:
- Benjamin Jean-Marie Tremblay^[benjamin.tremblay@uwaterloo.ca]
bibliography: universalmotif.bib
abstract: >
  This vignette walks through a complete ChIP-seq motif analysis on
  real data. Starting from a peak BED file bundled in
  `inst/extdata`, we extract peak sequences from a BSgenome,
  discover motifs de novo with `motif_finder()`, deduplicate them
  with `merge_similar2()`, compare them against the JASPAR2022
  plant collection via `MotifDb`, validate enrichment against a
  composition-matched background with `match_bkg()` and
  `enrich_motifs2()`, scan all peaks with `scan_sequences2()`,
  test for central positional bias with `motif_peaks()`, and look
  for tight motif self-clustering with `motif_coocc()`. The
  closing section assembles a summary table that combines all of
  the per-motif evidence into one ranked list.
vignette: >
  %\VignetteIndexEntry{An end-to-end ChIP-seq workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  bookdown::pdf_document2
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages({
  library(universalmotif)
  library(Biostrings)
  library(GenomicRanges)
})
options(Biostrings.coloring = FALSE)

has_pkgs <- suppressWarnings(suppressMessages(vapply(
  c("rtracklayer", "BSgenome.Athaliana.TAIR.TAIR9", "MotifDb"),
  requireNamespace, logical(1), quietly = TRUE)))
if (!all(has_pkgs)) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Skipping ChIP-seq vignette: required Suggests packages ",
          "are not installed (",
          paste(names(has_pkgs)[!has_pkgs], collapse = ", "),
          ").")
}
```

# Introduction

This vignette ties together the analysis functions of the
`universalmotif` package into a single end-to-end ChIP-seq workflow.
The previous vignettes cover individual pieces in isolation: for a
introduction to sequence motifs, see the
[introductory](IntroductionToSequenceMotifs.pdf) vignette; for the
plumbing of motif objects (I/O, conversion, logos), see the
[motif manipulation](MotifManipulation.pdf) vignette; and for the
mechanics of scanning, shuffling, and enrichment, see the
[sequence searches](SequenceSearches.pdf) vignette. This document
assumes you have skimmed the first and last of those.

The dataset is a real ChIP-seq peak call set for the *Arabidopsis
thaliana* B3-domain repressor VAL1, taken from @yuan2021val. VAL1
and its paralogue VAL2 silence seed-maturation genes by binding the
RY element (`CATGCA` / `TGCATG`) and recruiting Polycomb Repressive
Complex 2. We expect a properly working motif-discovery pipeline,
fed only the peaks, to recover the RY element and to match it
against B3-domain transcription factors in JASPAR.

The workflow proceeds in eight steps:

1. Import the peak BED.
2. Extract peak sequences from a BSgenome.
3. Discover motifs *de novo* with `motif_finder()`.
4. Deduplicate the discovered set with `merge_similar2()`.
5. Compare the survivors to JASPAR via `compare_motifs2()`.
6. Validate enrichment against a composition-matched background
   with `match_bkg()` and `enrich_motifs2()`.
7. Map hit positions with `scan_sequences2()` and test for central
   bias with `motif_peaks()`.
8. Look for tight motif self-clustering with `motif_coocc()`.

We close with a summary table that ranks the discovered motifs by
the combined evidence.

# Loading peaks from a BED file

`universalmotif` does not ship its own BED importer because
`rtracklayer::import()` already does the job well. The peak file
lives in `inst/extdata`; `system.file()` resolves it to an absolute
path inside the installed package.

```{r}
library(rtracklayer)

bed.file <- system.file("extdata",
                        "VAL1-GFP_val1_calledPeaks.bed",
                        package = "universalmotif")
peaks <- import(bed.file, format = "BED")
peaks
```

The BED file uses bare numeric chromosome names (`"1"` through
`"5"`), whereas the matching BSgenome we will use in the next
section uses the `"Chr1"`..`"Chr5"` convention. A single
`seqlevels()` assignment fixes that, and we drop any mitochondrial
or plastid contigs at the same time (there are none in this peak
set, but it is a useful defensive habit).

```{r}
seqlevels(peaks) <- paste0("Chr", seqlevels(peaks))
peaks <- keepSeqlevels(peaks,
                       paste0("Chr", 1:5),
                       pruning.mode = "coarse")
summary(width(peaks))
length(peaks)
```

# Extracting peak sequences and building a matched background

We use the TAIR9 BSgenome and resize every peak to a fixed 500 bp
window centred on the called peak. The fixed window matters
because `motif_peaks()` later treats positions as comparable across
sequences; if the peaks varied in length, the "centre" would mean
different things in different rows.

```{r}
library(BSgenome.Athaliana.TAIR.TAIR9)

peaks.500 <- resize(peaks, width = 500, fix = "center")
## Drop any windows that ran off a chromosome end.
peaks.500 <- trim(peaks.500)
peaks.500 <- peaks.500[width(peaks.500) == 500]

peak.seqs <- getSeq(Athaliana, peaks.500)
names(peak.seqs) <- paste0("peak_", seq_along(peak.seqs))
peak.seqs
```

Plant promoter sequences are strongly AT-skewed, so any background
analysis on these peaks needs to control for composition. We build
that control once, here, and reuse it for both *de novo* discovery
in §4 and the enrichment validation in §7. `match_bkg()` accepts a
`BSgenome` directly: pass `genome = Athaliana` and `exclude =
peaks.500`, and the function samples random non-peak genomic
windows internally and matches each peak on GC content and length.

```{r}
set.seed(2026)
bkg.seqs <- match_bkg(peak.seqs, genome = Athaliana, exclude = peaks.500)
length(bkg.seqs)
```

Sanity-check that the peak set and the matched background have
indistinguishable GC distributions:

```{r}
summary(letterFrequency(peak.seqs, "GC", as.prob = TRUE))
summary(letterFrequency(bkg.seqs,  "GC", as.prob = TRUE))
```

# De novo motif discovery

`motif_finder()` is a self-contained k-mer enumeration discoverer
that does not require an external MEME install. By default it
generates its own negative pool by shuffling the input at
`shuffle.k = 2`; passing `bkg.sequences = bkg.seqs` (our
composition-matched genomic pool from §3) is meaningfully better
on plant promoter ChIP-seq, because shuffling each peak preserves
the peak's own AT-skew and so AT-rich k-mers can't enrich, which
hides everything past the most-extreme signal. With a real
genomic background, lower-abundance motifs (G-box, E-box)
come out alongside the dominant RY element. On this
dataset (5,318 peaks of 500 bp each, two threads) the call returns
in roughly 15 seconds, so there is no need to sub-sample.

```{r motif-finder, cache=FALSE}
discovered.df <- motif_finder(peak.seqs,
                              bkg.sequences = bkg.seqs,
                              nmotifs       = 5,
                              min.width     = 6,
                              max.width     = 12,
                              rng.seed      = 2026,
                              nthreads      = 2)
print(as.data.frame(discovered.df[, c("name", "consensus", "pval")]),
      row.names = FALSE)
```

`motif_finder()` returns a `universalmotif_df` whose `motif` column
wraps the actual motif objects in an `AsIs` envelope. `to_list()`
unwraps it back into a plain list, which is what every downstream
function expects.

```{r}
discovered <- to_list(discovered.df, extrainfo = FALSE)
```

For viewing, we will create a copy which has the discovery
p-value added to the name:

```{r, fig.height=7, fig.width=7}
discovered.labelled <- discovered
for (i in seq_along(discovered.labelled)) {
  discovered.labelled[[i]]["name"] <- sprintf(
    "%s  (p = %.1e)",
    discovered.labelled[[i]]["name"],
    discovered.df$pval[i]
  )
}
view_motifs(discovered.labelled, sort.by = "similarity",
            show.positions = FALSE, names.pos = "right")
```

With `sort.by = "similarity"`, `view_motifs()` reorders the panels
by hierarchical clustering on Pearson correlation between aligned
columns, so visually related logos are adjacent. The RY-element
variants (consensuses containing `CATGCA`) cluster together; so do
the G-box (`CACGTG`) variants, 
GAGA repeats, and low-complexity AT-rich runs. The next two sections
(enrichment against a composition-matched background, and
positional bias) confirm which of these clusters carry real
binding signal versus background composition. Pass
`sort.by = "ic"` (the default) for descending information content,
or `sort.by = "none"` to keep `motif_finder()`'s p-value ranking.

# Deduplicating the discovered set

A real motif-discovery run almost always returns several rotated or
extended copies of the same underlying signal. `merge_similar2()`
collapses them into a single representative per cluster by
hierarchical clustering on q-values from a Pearson-based comparison.
It's often wise to also trim the resulting merged motifs, since
merging often collapses mismatching flanks into low IC positions.

```{r}
discovered.merged <- merge_similar2(discovered, qvalue = 0.01)
discovered.merged <- trim_motifs(discovered.merged)

length(discovered)
length(discovered.merged)
```

`merge_similar2()` names merged motifs by joining the names of the
contributing motifs with `+`, which gets unwieldy fast when several
distinct binding signals each produce their own cluster (the RY
element, the bZIP G-box, and so on). Relabel each merged cluster
by inspecting its consensus, so the rest of the vignette can refer
to motifs by what they *are*:

```{r}
label_merged <- function(consensus) {
  if (grepl("CATGCA", consensus))                    "RY"
  else if (grepl("CACGTG", consensus))               "G_box"
  else if (grepl("ACGTAC|GTACGT", consensus))        "bZIP"
  else if (grepl("AAAAAA|ATATAT|AAATAA", consensus)) "AT_rich"
  else if (grepl("GAGAGA", consensus))               "GA_repeat"
  else                                               consensus
}
new.names <- vapply(discovered.merged,
                    function(m) m["name"], character(1))
for (i in seq_along(discovered.merged)) {
  new.names[i] <- label_merged(
    discovered.merged[[i]]["consensus"]
  )
  discovered.merged[[i]]["name"] <- new.names[i]
}
new.names
```

```{r, fig.height=7, fig.width=7}
view_motifs(discovered.merged, sort.by = "similarity",
            show.positions = FALSE, names.pos = "right")
```

# Comparing against JASPAR via MotifDb

For the cross-database comparison we use `MotifDb`, which already
ships JASPAR2022 plant motifs and is therefore the lightest path to
a reference set. We narrow it to *Arabidopsis* JASPAR2022 entries
and convert the resulting list to `universalmotif` objects.

```{r motifdb, warning=FALSE, message=FALSE}
suppressPackageStartupMessages(library(MotifDb))

ath.jaspar <- query(query(MotifDb, "Athaliana"), "jaspar2022")
length(ath.jaspar)
ath.motifs <- convert_motifs(ath.jaspar)
```

`compare_motifs2()` in long-format mode treats indexed entries of
`motifs` as queries and the whole input as the target database, so
we concatenate the discovered set with the reference and pass the
query indices via `compare.to`.

```{r}
combined <- c(discovered.merged, ath.motifs)
n.disc   <- length(discovered.merged)

cmp <- compare_motifs2(combined,
                       compare.to = seq_len(n.disc),
                       qvalue     = 0.2,
                       nthreads   = 2)
cmp <- cmp[cmp$subject != cmp$target, ]
cmp <- cmp[order(cmp$qvalue), ]
print(head(cmp[, c("subject", "target", "score", "qvalue",
                   "subject.consensus", "target.consensus")], 10),
      row.names = FALSE)
```

Most of the strong matches will be for AT-rich or GA-rich
low-complexity discovered motifs (DOF, BPC, and so on) rather than
for the RY element itself; the *Arabidopsis* genome is AT-skewed
and `motif_finder()` reports every strong k-mer signal, not only
the one that drove the immunoprecipitation. The next two sections
(enrichment against a composition-matched background, and
positional bias) are what separate the binding signal from
incidental low-complexity hits.

For the visual, pick the RY motif and pair it with its best JASPAR
match in the B3-domain family. ABI3, FUS3, and LEC2 are the three
B3-domain TFs in JASPAR2022 for *Arabidopsis*, and any of them is a
biologically sensible neighbour for VAL1.

```{r, fig.height=2.5, fig.width=7}
ref.names <- vapply(ath.motifs, function(m) m["name"], character(1))
ry.cmp <- cmp[cmp$subject == "RY", ]
ry.best <- ry.cmp[order(ry.cmp$qvalue), ][1, ]

top.disc <- discovered.merged[[match("RY", new.names)]]
top.ref  <- ath.motifs[[match(ry.best$target, ref.names)]]
view_motifs(c(top.disc, top.ref), show.positions.once = FALSE,
            names.pos = "right")
```

# Validating enrichment against a composition-matched background

We can now use `enrich_motifs2()` to validate our new set of merged motifs.
`enrich_motifs2()` accepts an explicit `bkg.sequences` argument so
we can supply a background that controls for the AT-skew of plant
promoters. We already built that pool in §3 (`bkg.seqs`, from
`match_bkg()` against a universe of random non-peak genomic
windows), so the enrichment step just reuses it.

Feed the matched background into `enrich_motifs2()`:

```{r}
enr <- enrich_motifs2(motifs        = discovered.merged,
                      sequences     = peak.seqs,
                      bkg.sequences = bkg.seqs,
                      pvalue        = 1e-4,
                      qvalue        = 0.05,
                      nthreads      = 2)
print(enr[, c("motif", "consensus", "enrichment", "qvalue")],
      row.names = FALSE)
```

With a GC-matched background, the AT-rich low-complexity motifs
drop to barely-above-1 fold enrichment (because they are common in
*every* AT-skewed plant promoter window), while other motifs stands
out at several fold over background. That contrast is invisible if
you compare against a uniform shuffled null, which is exactly why
this kind of matched background matters for plant ChIP-seq.

For the alternative dinucleotide-shuffle null (`shuffle.k = 2`), or
for a single-shuffle background, see the `match_bkg()` and
shuffling sections of the
[sequence searches](SequenceSearches.pdf) vignette.

# Scanning all peaks and testing positional bias

To map hits at single-base resolution across every peak, use
`scan_sequences2()` with the discovered motifs. The 500 bp fixed
window means motif positions are directly comparable across rows.

```{r}
hits <- scan_sequences2(discovered.merged,
                        peak.seqs,
                        pvalue         = 1e-4,
                        return.granges = FALSE,
                        nthreads       = 2)
nrow(hits)
table(motif = hits$motif)
```

`motif_peaks()` then asks, per motif, whether the hits are clustered
near the peak centre. It is a non-parametric CentriMo-style test:
for each window size, the function compares the observed hit count
inside the window to the count expected under a uniform-position
null.

```{r}
pk <- motif_peaks(hits, seq.length = 500, mode = "central")
print(pk[, c("motif", "best.center", "best.window", "qvalue")],
      row.names = FALSE)
```

```{r, fig.height=3.5, fig.width=7}
plot_motif_peaks(pk)
```

A motif that drives binding should peak sharply at position 250,
the centre of the 500 bp window. The RY element does (very
strongly, since VAL1 is the immunoprecipitated factor); the G-box
/ bZIP motifs are also centrally enriched but with a
broader best window, consistent with bZIP TFs co-binding nearby
without being the primary signal. The low-complexity AT-rich
motifs are filtered out by the `motif_peaks()` default
`qvalue = 0.1` threshold.

# Motif clustering with `motif_coocc()`

`motif_coocc()` asks whether motif pairs appear together in the
same peak (and, optionally, within `max.distance` bp of each other)
more often than expected by chance. For VAL1 the interesting
question is not so much "which pair?" as "does the RY element
cluster with itself inside peaks?". B3-domain TFs often bind
cooperatively to tandem RY elements, and ChIP-seq peaks are wide
enough to accommodate several copies. We turn on `self.pairs` so
the (motif, motif) self-pairs are included:

```{r}
co <- motif_coocc(discovered.merged,
                  hits         = hits,
                  n.sequences  = length(peak.seqs),
                  max.distance = 50L,
                  self.pairs   = TRUE)
print(co[, c("motif_a", "motif_b", "both", "both.clustered",
             "median.distance", "qvalue")],
      row.names = FALSE)
```

The (`RY`, `RY`) self-pair shows how often the RY element appears
in tight clusters within a peak (`both.clustered` counts sequences
where two RY hits land within 50 bp of each other, with
`median.distance` giving the typical spacing). A high
`both.clustered` for the RY self-pair, with a small
`median.distance`, is the cooperative-binding signature: multiple
RY copies packed together at the binding locus. The cross-motif
rows give the rest of the picture: pairwise co-occurrence of the
RY motif with each low-complexity neighbour is generally weak (no
obligate cofactor), and the strongest cross-pairs are between the
low-complexity motifs themselves, reflecting plant promoter
composition rather than specific cobinding.

`motif_coocc()` reports the summary statistic, but it is worth
plotting the actual distribution: for every peak that carries two
or more RY hits, compute every pairwise distance between RY hit
starts and histogram them.

```{r, fig.height=3, fig.width=7}
ry.hits  <- hits[hits$motif == "RY", ]
ry.dists <- unlist(lapply(split(ry.hits$start, ry.hits$sequence.i),
                          function(s) if (length(s) < 2L)
                            integer(0) else as.integer(dist(s))))
length(ry.dists)
ggplot2::ggplot(data.frame(distance = ry.dists),
                ggplot2::aes(x = .data$distance)) +
  ggplot2::geom_histogram(binwidth = 5, fill = "black", colour = NA) +
  ggplot2::geom_vline(xintercept = 50, linetype = "dashed",
                      colour = "grey50") +
  ggplot2::labs(x = "RY-RY distance (bp)", y = "Count") +
  ggplot2::theme_bw() +
  ggplot2::theme(panel.grid         = ggplot2::element_blank(),
                 panel.border       = ggplot2::element_blank(),
                 axis.line.x.bottom = ggplot2::element_line(colour = "black"),
                 axis.line.y.left   = ggplot2::element_line(colour = "black"))
```

The dashed line marks the 50 bp threshold used in `max.distance`
above. A heavy tail at very short spacings (under 20 bp) is the
direct signature of cooperative B3-domain binding to tandem RY
elements.

The cross-motif case repeats the same calculation, this time
measuring the distance from each RY hit to every hit of every
other motif in the same peak. A pile-up at short distance would
indicate a TF whose binding co-localises with RY; a flat
distribution indicates incidental co-occurrence driven only by
both motifs being common in the peak set.

```{r, fig.height=5, fig.width=7}
ry.by.seq <- split(ry.hits$start, ry.hits$sequence.i)
other.motifs <- setdiff(unique(hits$motif), "RY")
cross.dists <- do.call(rbind, lapply(other.motifs, function(om) {
  om.hits   <- hits[hits$motif == om, ]
  om.by.seq <- split(om.hits$start, om.hits$sequence.i)
  shared    <- intersect(names(ry.by.seq), names(om.by.seq))
  d <- unlist(lapply(shared, function(s)
    abs(as.integer(outer(ry.by.seq[[s]], om.by.seq[[s]], "-")))))
  if (!length(d)) return(NULL)
  data.frame(other = om, distance = d, stringsAsFactors = FALSE)
}))
ggplot2::ggplot(cross.dists, ggplot2::aes(x = .data$distance)) +
  ggplot2::geom_histogram(binwidth = 10, fill = "black", colour = NA) +
  ggplot2::geom_vline(xintercept = 50, linetype = "dashed",
                      colour = "grey50") +
  ggplot2::facet_wrap(~ other, scales = "free_y") +
  ggplot2::labs(x = "RY-to-other distance (bp)", y = "Count") +
  ggplot2::theme_bw() +
  ggplot2::theme(panel.grid         = ggplot2::element_blank(),
                 panel.border       = ggplot2::element_blank(),
                 strip.background   = ggplot2::element_rect(fill = NA,
                                                            colour = NA),
                 axis.line.x.bottom = ggplot2::element_line(colour = "black"),
                 axis.line.y.left   = ggplot2::element_line(colour = "black"))
```

None of the cross-motif panels shows the dramatic lone 0-5 bp spike
that the (RY, RY) self-pair did. The AT-rich and GA-repeat panels
show slight increased abundance in the 50 bp window (indicating a possible
relationship between RY motifs and structural elements), but also many hits
outside of the window. The G_box and
bZIP panels do show a concentration of overall hits at short distance in
comparison, hinting at real colocalisation between RY and bZIP-family
motifs, which is biologically plausible for seed-maturation
regulatory regions where RY and G-box elements are known to
co-occur. The take-home is that RY's strongest self-pair signal is
its own self-clustering, with a secondary, much milder bZIP / G-box
neighbourship.

# Summary table

The final step is a compact ranked summary that pulls one row per
discovered motif from each of the per-motif analyses (best JASPAR
match, enrichment q, positional q) and adds a "best partner" (not self)
column from the co-occurrence table.

```{r}
disc.names <- new.names

best.match <- vapply(disc.names, function(nm) {
  i <- which(cmp$subject == nm)
  if (!length(i)) NA_character_ else cmp$target[i[1]]
}, character(1))

enrich.q <- vapply(disc.names, function(nm) {
  i <- which(enr$motif == nm)
  if (!length(i)) NA_real_ else enr$qvalue[i[1]]
}, numeric(1))

pos.q <- vapply(disc.names, function(nm) {
  i <- which(pk$motif == nm)
  if (!length(i)) NA_real_ else pk$qvalue[i[1]]
}, numeric(1))

best.partner <- vapply(disc.names, function(nm) {
  ix <- which((co$motif_a == nm | co$motif_b == nm) &
              co$motif_a != co$motif_b)
  if (!length(ix)) return(NA_character_)
  ix <- ix[which.min(co$qvalue[ix])]
  if (co$motif_a[ix] == nm) co$motif_b[ix] else co$motif_a[ix]
}, character(1))

summary.tbl <- data.frame(
  motif         = disc.names,
  jaspar.match  = best.match,
  enrich.q      = signif(enrich.q, 3),
  positional.q  = signif(pos.q, 3),
  best.partner  = best.partner,
  stringsAsFactors = FALSE
)
rownames(summary.tbl) <- NULL
print(summary.tbl, row.names = FALSE)
```

```{r, fig.height=7, fig.width=7}
view_motifs(discovered.merged, sort.by = "similarity",
            show.positions = FALSE, names.pos = "right")
```

Looking at the combined evidence, our analysis has clearly
recovered the dominant VAL1 RY motif, with a couple of secondary
motifs which may be binding sites of co-regulating TFs. The remaining
repetitive motifs, while significantly enriched, fail the
positional test and are likely incidental structural features of the
types of sequences VAL1 binds.

# Session info

```{r}
sessionInfo()
```

# References
