---
title: "Introduction to \"universalmotif\""
shorttitle: "Intro to \"universalmotif\""
date: 5 May 2026
author:
- Benjamin Jean-Marie Tremblay^[benjamin.tremblay@uwaterloo.ca]
bibliography: universalmotif.bib
vignette: >
  %\VignetteIndexEntry{Introduction to "universalmotif"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  bookdown::pdf_document2:
    toc: false
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>")
```

# Installation {.unnumbered}

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("universalmotif")
```

# Overview {.unnumbered}

For a brief explanation of sequence motifs, see the [Introduction to sequence motifs](IntroductionToSequenceMotifs.pdf) vignette. This broadly covers the different 'types' of motif representation, and establishes the nomenclature used by the `universalmotif` package.

The capabilities of the `universalmotif` package can be divided into several general categories. These are briefly demonstrated in the following vignettes:

* [Motif import, export, and manipulation](MotifManipulation.pdf)
* [Sequence scanning and manipulation](SequenceSearches.pdf)
* [Motif comparisons and P-values](MotifComparisonAndPvalues.pdf)

# Choosing a function {.unnumbered}

For ordinary DNA/RNA workflows, the `_lite` functions provide a smaller
interface. The original functions remain necessary for features outside that
interface; `_lite` does not mean a drop-in replacement for every argument or
output type.

Task | DNA/RNA starting point | Use the original function when you need
-----|-----------------------|-----------------------------------------
Scan sequences | `scan_sequences_lite()` | Amino-acid/custom alphabets, higher-order or gapped motifs, non-P-value thresholds, exhaustive P-values, or q-values
Compare motifs | `compare_motifs_lite()` | Other alphabets, non-PCC metrics, higher-order comparisons, IC filters, or alternative score aggregation
Enrichment | `enrich_motifs_lite()` | Scanning options outside the lite interface; check the counting and statistical-test defaults before switching
Align logos or draw trees | `view_motifs_lite()`, `motif_tree_lite()` | Other alphabets or comparison options available only in the original functions
Merge motifs | `merge_motifs_lite()`, `merge_similar_lite()` | Other alphabets or the broader original comparison and clustering options

`motif_pvalue()` is shared by both scanner interfaces. Its dynamic method uses
an integerised score distribution; see the P-value vignette for inclusive
thresholds, unattainable cutoffs, and background rebasing.

CWMs can be plotted explicitly with `use.type = "CWM"`; they are not
probability matrices or ordinary scanning PWMs. See the contribution-weight
section of the sequence-motif introduction before converting them.

Use `saveRDS()`/`readRDS()` for lossless storage of complete motif objects.
The native `write_motifs()` format retains gap definitions and, by default,
higher-order matrices, but rounds matrix entries. External motif formats
preserve different subsets of metadata and warn when gaps or higher-order
matrices cannot be represented.
