---
title: "Introduction to sequence motifs"
shorttitle: "Intro to motifs"
author:
- name: Benjamin Jean-Marie Tremblay
  affiliation: University of Waterloo, Waterloo, Canada
  email: b2tremblay@uwaterloo.ca
bibliography: universalmotif.bib
vignette: >
  %\VignetteIndexEntry{Introduction to sequence motifs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::pdf_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>")
suppressPackageStartupMessages(library(universalmotif))
```

# Introduction to sequence motifs

Motifs are a more practical representation of consensus elements in biological
sequences, allowing for a more detailed description of the variability at each
site [see @motifhist]. There are a number of ways of representing motifs; four
types are available here. All of these can be stored using the same
`universalmotif` class. Unfortunately, the naming conventions for the various
motif types are not unanimous. In order to provide a simple interface to the
various types, in this package they are referred to as Position Count Matrix
(PCM), Position Probability Matrix (PPM), Position Weight Matrix (PWM), and
Information Content Matrix (ICM). 

These four types will be described here. Furthermore, the actual code used to
go between these types can be see in `?utilities`.

# Position count matrices

Also known as position frequency matrices, these are typically the most basic
representation of a motif. Simply, for each position the total counts of
letters is shown. For example, the following sequences:

 # | Sequence
-- | --------
 1 | AAGAAT
 2 | ATCATA
 3 | AAGTAA
 4 | AACAAA
 5 | ATTAAA
 6 | AAGAAT
: (\#tab:seqs) Starting sequences.

... would be represented as:

Position | 1 | 2 | 3 | 4 | 5 | 6
-------- | - | - | - | - | - | -
       A | 6 | 4 | 0 | 5 | 5 | 4
       C | 0 | 0 | 2 | 0 | 0 | 0
       G | 0 | 0 | 3 | 0 | 0 | 0
       T | 0 | 2 | 1 | 1 | 1 | 2
: (\#tab:pcm) Position count matrix.

_(Note that all positions sum to 6, the initial number of sequences.)_

# Position probability matrices

Also sometimes known as position frequency matrices and position weight
matrices, these represent the probabilities for each letter at each position.
Using the previous motif (table \@ref(tab:pcm)) as an example, the following
formula would be used to calculate the probability of each letter _N_
at each individual position:

\begin{equation}
  Prob_N = \frac{Counts_N}{Counts_{Total}}
  (\#eq:pcm2ppm)
\end{equation}

This leads to the following motif representation:

Position |    1 |    2 |    3 |    4 |    5 |    6
-------- | ---- | ---- | ---- | ---- | ---- | ----
       A | 1.00 | 0.67 | 0.00 | 0.83 | 0.83 | 0.66
       C | 0.00 | 0.00 | 0.33 | 0.00 | 0.00 | 0.00
       G | 0.00 | 0.00 | 0.50 | 0.00 | 0.00 | 0.00
       T | 0.00 | 0.33 | 0.17 | 0.17 | 0.17 | 0.33
: (\#tab:ppm) Position probability matrix.

_(Note that all positions sum to 1.)_

From this type of representation, the probability of any combination of letters
can be calculated. For example, the probability for AAGAAA is about 15%.
However, when starting from a small pool of sequences, many zeroes can appear
in the PPM; meaning that, for example, the probability of AAAAAA is currently
zero. When scanning through large numbers of biological sequences, throwing away
combinations of
letters such as these can be undesirable, as for example  mimatches can be quite
common in transcription factor binding sites; this can be fixed by adding a
'pseudocount'. Usually a small number such as 1, it is introduced into the PCM
to PPM caculation:

\begin{equation}
  Prob_N = \frac{Counts_N + \frac{Pseudocount}{length_N}}{Counts_{Total} + Pseudocount}
  (\#eq:ppmpseudo)
\end{equation}

In this equation, the pseudocount is added to the top and bottom of the
fraction. However for the top fraction, which is specific to each
letter, the pseudocount is divided by the total number of letters (in the case
of DNA, 4). This then generates the following motif:

Position |     1 |     2 |     3 |     4 |     5 |     6
-------- | ----- | ----- | ----- | ----- | ----- | -----
       A | 0.892 | 0.610 | 0.036 | 0.750 | 0.750 | 0.610
       C | 0.036 | 0.035 | 0.320 | 0.035 | 0.035 | 0.035
       G | 0.036 | 0.035 | 0.464 | 0.035 | 0.035 | 0.035
       T | 0.036 | 0.320 | 0.180 | 0.180 | 0.180 | 0.320
: (\#tab:ppmpseudo) Position probability matrix with a pseudocount of 1.

Now, though unlikely, it is no longer considered impossible for the sequence
AAAAAA to exist as part of this motif. Since the total number of sequences in
this case is quite low, the pseudocount can have a large impact on the values
within the matrix; this impact decreases as the number of sequences increases.
Of course, this can also be changed by using different pseudocounts.

# Position weight matrices

The position weight matrix, also known as position-specific weight matrix,
position-specific scoring matrix, and logodds scoring matrix, was first
proposed by @pwm. In this case for
each position, every letter has a 'score'; this can be used to evaluate how
well a sequence matches a motif. Though there can be multiple ways of calculating
these scores, the most common method is to calculate the log of each
probability, correcting for background frequencies. This results in the
following calculation:

\begin{equation}
  Score_N = \log_{2}\left(\frac{Prob_N}{ProbBkg_N}\right)
  (\#eq:ppm2pwm)
\end{equation}

Using this equation, the log of fractions where the probability of a
certain letter in
a sequence is higher than that of the background probability of that letter
result in positive scores, and vice versa for negative scores. Using the table
\@ref(tab:ppm) motif and assuming a uniform background frequency (i.e. the
probability of each of the four letters is 0.25), this results in the following PWM:

Position |    1 |     2 |      3 |      4 |      5 |     6
-------- | ---- | ----- | ------ | ------ | ------ | -----
       A |    2 | 1.425 |   -Inf |  1.737 |  1.737 | 1.415
       C | -Inf |  -Inf |  0.415 |   -Inf |   -Inf |  -Inf
       G | -Inf |  -Inf |  1.000 |   -Inf |   -Inf |  -Inf
       T | -Inf | 0.415 | -0.585 | -0.585 | -0.595 | 0.415
: (\#tab:pwm) Position weight matrix.

_(Note that the position totals no longer have equal sums.)_

In order to score a sequence, add up the score for the letters at the specific
positions; for example AAGAAA has a score of 9.31. However, similar to with
PPMs, if starting from a small pool of sequences the sequence AAAAAA could
never be recovered using this motif model, with a score of -Inf. This can
be avoided simply by starting from a pseudocount-adjusted PPM. Using the table
\@ref(tab:ppmpseudo) motif, this becomes:

Position |      1 |      2 |      3 |      4 |      5 |      6
-------- | ------ | ------ | ------ | ------ | ------ | ------
       A |  1.840 |  1.280 | -2.807 |  1.585 |  1.585 |  1.280
       C | -2.807 | -2.807 |  0.363 | -2.807 | -2.807 | -2.807
       G | -2.807 | -2.807 |  0.893 | -2.807 | -2.807 | -2.807
       T | -2.807 |  0.363 | -0.485 | -0.485 | -0.485 |  0.363
: (\#tab:pwmpseudo) Position weight matrix with a pseudocount of 1.

Now, the score for AAGAA is 8.46, and the score for AAAAAA is 4.76. Though the
score for the latter is low, it is no longer -Inf as a result of one mismatch.

When searching for instances of this motif using this scoring system, one more
consideration is needed: a minimum score. For example, both sequences AAGAAA
and ATTTTT have positive scores; but one is much higher than the other. Should
ATTTTT then be discarded? In order to answer this a threshold is set; typically
this is a certain percent of the highest possible score. For example, in the table
\@ref(tab:pwmpseudo) motif, the highest possible score is 8.46; using
a threshold of 25%, the minimum score then becomes 2.115. This means that
ATTTTT, with a score of 1.112, would indeed be discarded.

Other methods for determining the minimum score include starting from
P-values. The package `r CRANpkg("TFMPvalue")` for example can calculate minimum
scores from P-values, using the algorithm described by @pval. The
`r Biocpkg("universalmotif")` package also offers this capability with the
`motif_pvalue()` function.

# Information content matrices

Finally, the information content matrix [@correction; @icm]. This type aims to
include another
consideration: are some positions more important than others? To explore this,
let us consider the table \@ref(tab:ppm) motif. This matrix can be represented
as a probability sequence logo as seen in figure \@ref(fig:logo1).

```{r logo1,fig.cap="Sequence logo of a PPM",echo=FALSE,fig.small=TRUE}
motif <- create_motif(c("AAGAAT", "ATCATA", "AAGTAA", "AACAAA", "ATTAAA",
                        "AAGAAT"), type = "PPM", pseudocount = 0)
view_motifs(motif, use.type="PPM")
```

In this case, the height of the letters represent their probabilities at each
position. However, when represented as an ICM, the sequence logo then
resembles as seen in figure \@ref(fig:logo2).

```{r logo2,fig.cap="Sequence logo of an ICM",echo=FALSE,fig.small=TRUE}
motif <- create_motif(c("AAGAAT", "ATCATA", "AAGTAA", "AACAAA", "ATTAAA",
                        "AAGAAT"), type = "PPM", pseudocount = 0)
view_motifs(motif)
```

Now, the total height of each position is scaled using the total
_information_ at that position. Simply put, the total information of each
position is an indication of the level of conservation; in the example motif,
the first position is highly conserved, always being the letter A, whereas the
third position is less so conserved, as the probabilities for any one letter
are quite lower.

For every position, the letter heights are calculated as the total possible
information content (IC) or maximum uncertainty, then subtracting that
positions' actual uncertainty
[@correction; @icm]. This calculation is based on Shannon's entropy
[@shannon], with the final values representing 'bits' [@bits].
The total IC is based on alphabet length,
using the following equation:

\begin{equation}
  IC_{total} = \log_{2}(length_N)
  (\#eq:totalic)
\end{equation}

For DNA motifs which have an alphabet length of 4, the total IC is 2. To
calculate the positions' actual uncertainty, the following equation is used:

\begin{equation}
  Uncertainty = - \sum_{N=A}^{T} Prob_N \times{} \log_{2}(Prob_N)
  (\#eq:posuncer)
\end{equation}

Where the position uncertainty is the sum of the uncertainty of all alphabet
letters (A, C, G, and T). To calculate the final information content:

\begin{equation}
  IC_{position} = IC_{total} - Uncertainty
  (\#eq:ic)
\end{equation}

In the original implementation described by @correction, an additional error
correction factor is included to account for sample size. This correction is
rarely used however, but can be incorporated as such:

\begin{equation}
  IC_{position} = IC_{total} - [ Uncertainty + E_{corr} ]
  (\#eq:iccorr)
\end{equation}

The details for calculating this factor will not be covered here; refer to
@correction. The `r Biocpkg("TFBSTools")` package offers the ability to
incorporate this error correction (and used in
the `r Biocpkg("universalmotif")` package via `convert_type()`).

Finally, to get the height of each letter at each position, the final IC is
multiplied by the letter and position probabilities:

\begin{equation}
  IC_N = Prob_N \times{} IC_{position}
  (\#eq:finalic)
\end{equation}

Using the above equations, the table \@ref(tab:ppm) motif then becomes:

Position |      1 |      2 |      3 |      4 |      5 |      6
-------- | ------ | ------ | ------ | ------ | ------ | ------
       A |  2.000 |  0.721 |  0.000 |  1.125 |  1.125 |  0.721
       C |  0.000 |  0.000 |  0.180 |  0.000 |  0.000 |  0.000
       G |  0.000 |  0.000 |  0.270 |  0.000 |  0.000 |  0.000
       T |  0.000 |  0.361 |  0.090 |  0.225 |  0.225 |  0.361
: (\#tab:icm) Information content matrix.

_(Note that none of the positions have a sum larger than 2.)_

An alternative to representing information content as Shannon's entropy is
relative entropy, or Kullback-Leibler divergence [@kl]. While IC as Shannon's
entropy has the advantage of having a consistent maximum IC for every position,
it does not take into account non-uniform background frequencies. Relative
entropy on the other hand will take this into account, but all positions no
longer share the same maximum IC. To calculate relative entropy:

\begin{equation}
  IC_N = Prob_N \times{} \log_{2}\left(\frac{Prob_N}{ProbBkg_N}\right)
  (\#eq:kldiv)
\end{equation}

Using this equation can lead to IC less than zero; these values are not
allowed, so they are simply replaced with zero.
With this equation and assuming uniform background frequencies, the table
\@ref(tab:ppm) motif becomes:

Position |      1 |      2 |      3 |      4 |      5 |      6
-------- | ------ | ------ | ------ | ------ | ------ | ------
       A |  1.640 |  0.777 |  0.000 |  1.190 |  1.190 |  0.777
       C |  0.000 |  0.000 |  0.177 |  0.000 |  0.000 |  0.000
       G |  0.000 |  0.000 |  0.415 |  0.000 |  0.000 |  0.000
       T |  0.000 |  0.177 |  0.090 |  0.000 |  0.000 |  0.117
: (\#tab:icmkl) Information content matrix as relative divergence.

This motif would look significantly different with non-uniform background
frequences. For example, starting from the following background frequencies: 
`c(A = 0.4, C = 0.1, G = 0.1, T= 0.4)`, the motif resembles:

Position |      1 |      2 |      3 |      4 |      5 |      6
-------- | ------ | ------ | ------ | ------ | ------ | ------
       A |  1.030 |  0.366 |  0.000 |  0.680 |  0.680 |  0.366
       C |  0.000 |  0.000 |  0.541 |  0.000 |  0.000 |  0.000
       G |  0.000 |  0.000 |  1.028 |  0.000 |  0.000 |  0.000
       T |  0.000 |  0.000 |  0.090 |  0.000 |  0.000 |  0.000
: (\#tab:icmkl2) Information content matrix as relative divergence with
a non-uniform background.

# References {.unnumbered}
