---
title: "Introduction to \"universalmotif\""
shorttitle: "Intro to \"universalmotif\""
author:
- name: Benjamin Jean-Marie Tremblay
  affiliation: University of Waterloo, Waterloo, Canada
  email: b2tremblay@uwaterloo.ca
bibliography: universalmotif.bib
vignette: >
  %\VignetteIndexEntry{Introduction to "universalmotif"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::pdf_document:
    latex_engine: lualatex
    toc: false
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>")
```

# Installation {.unnumbered}

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("universalmotif")
```

# Overview {.unnumbered}

For a brief explanation of sequence motifs, see the
[Introduction to sequence motifs](IntroductionToSequenceMotifs.pdf) vignette. This
broadly covers the different 'types' of motif representation, as well as
establishes the nomenclature used by the `r Biocpkg("universalmotif")` package.

Broadly speaking, the capabilities of the `r Biocpkg("universalmotif")` package
can be divided into three categories. These are briefly demonstrated in the
following vignettes:

* [Motif import, export, and manipulation](MotifManipulation.pdf)
* [Sequence scanning and manipulation](SequenceSearches.pdf)
* [Motif analysis and advanced usage](AdvancedUsage.pdf)

