---
title: "Building a curated motif database from multiple sources"
shorttitle: "Curated motif databases"
date: 25 May 2026
author:
- Benjamin Jean-Marie Tremblay^[benjamin.tremblay@uwaterloo.ca]
bibliography: universalmotif.bib
abstract: >
  In practice, a motif scan rarely relies on a single public
  database. This vignette walks through curating one consolidated
  motif set
  from several public sources, using *Homo sapiens* motifs pulled
  out of three `MotifDb` sub-collections (JASPAR2022,
  HOCOMOCOv11-core, CIS-BP) as the working example. We import,
  inspect cross-source redundancy with `compare_motifs2()`,
  deduplicate via graph-based clustering in `merge_similar2()`,
  trim low-information flanks with `trim_motifs()`, annotate the
  survivors with the existing metadata via `to_df()` /
  `update_motifs()`, and export the whole curated library as a
  single MEME file ready for downstream use.
vignette: >
  %\VignetteIndexEntry{Building a curated motif database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  bookdown::pdf_document2
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.pos = "H")
suppressPackageStartupMessages({
  library(universalmotif)
  library(ggplot2)
})

has_pkgs <- suppressWarnings(suppressMessages(vapply(
  c("MotifDb", "ape", "ggtree"),
  requireNamespace, logical(1), quietly = TRUE)))
if (!all(has_pkgs)) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Skipping curated-DB vignette: required Suggests packages ",
          "are not installed (",
          paste(names(has_pkgs)[!has_pkgs], collapse = ", "),
          ").")
}
```

# Introduction

No single public motif database covers every transcription factor
in every organism, and no single source is perfectly consistent: each
ships its own naming conventions, its own matrix-type conventions
(PCM, PPM, or PWM), and its own intra-source redundancies (several
matrices for the same factor, often with slight quality
differences). Anyone doing applied motif scanning sooner or later
ends up pulling from several sources and reconciling them more or
less by hand. This vignette gathers the building blocks that
`universalmotif` provides for that reconciliation into a single
end-to-end recipe, demonstrated here on a few hundred *Homo sapiens*
motifs drawn from three `MotifDb` sub-collections.

The previous vignettes cover the underlying processes in
isolation: for single-motif import and export, see the
[motif manipulation](MotifManipulation.pdf) vignette; for the
mechanics of motif-vs-motif scoring and p-value calculation, see
the [comparisons and p-values](MotifComparisonAndPvalues.pdf)
vignette; for an end-to-end ChIP-seq analysis that would use a
curated motif library, see the
[ChIP-seq workflow](ChIPseqWorkflow.pdf) vignette. This vignette
sits one step upstream of the ChIP-seq one: it produces the kind
of consolidated MEME file that a real analyst would feed into
`scan_sequences2()` or `enrich_motifs2()`.

The recipe proceeds in eight steps:

1. Pull motifs from three sources via `MotifDb`.
2. Mark the originating source onto each motif.
3. Inspect intra- and inter-source redundancy with
   `compare_motifs2()`.
4. Deduplicate via graph-based clustering with
   `merge_similar2()`.
5. Visualise the cluster structure with `motif_tree2()`.
6. Trim uninformative flanks with `trim_motifs()`.
7. Annotate the survivors with metadata via `to_df()` /
   `update_motifs()`.
8. Export the curated set as a single MEME file with
   `write_meme()`.

# Loading motifs from several sources

`MotifDb` ships a large indexed collection that has already been
parsed from a wide variety of source formats. For a single
demonstration we narrow it to *Homo sapiens* and pull three
sub-collections that, between them, cover a fair slice of
human TFs: `jaspar2022`, `HOCOMOCOv11-core-A`, and `cisbp_1.02`.

```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(MotifDb))

hs <- suppressWarnings(query(MotifDb, "hsapiens"))
sources <- c("jaspar2022", "HOCOMOCOv11-core-A", "cisbp_1.02")
per.source <- suppressWarnings(lapply(sources, function(s)
  convert_motifs(query(hs, s))))
names(per.source) <- sources
sapply(per.source, length)
```

We annotate each motif's `family` slot with the originating source.
Doing this here, before the lists are concatenated, means the
source survives every subsequent operation (clustering,
trimming, MEME export) and gives `motif_tree2()` a per-motif
colouring column to use later. `family` is the natural slot for
this because `MotifDb` rarely populates it, and because
`motif_tree2(linecol = "family")` is the function's default.

```{r}
for (s in sources) {
  for (i in seq_along(per.source[[s]]))
    per.source[[s]][[i]]["family"] <- s
}
all.motifs <- do.call(c, per.source)
length(all.motifs)
```

As a quick visual sanity check, we can plot the per-source counts:

```{r, fig.height=2.5, fig.width=6}
src.df <- data.frame(source = sources,
                     count  = sapply(per.source, length),
                     stringsAsFactors = FALSE)
ggplot(src.df, aes(x = .data$source, y = .data$count)) +
  geom_col(fill = "black") +
  geom_text(aes(label = .data$count), vjust = -0.4) +
  labs(x = NULL, y = "Number of motifs") +
  theme_bw() +
  theme(panel.grid         = element_blank(),
        panel.border       = element_blank(),
        axis.line.x.bottom = element_line(colour = "black"),
        axis.line.y.left   = element_line(colour = "black"))
```

## Sidebar: importing raw files directly

The rest of this vignette assumes the motifs were sourced through
`MotifDb`. If instead you have raw downloaded files sitting on disk,
the import step is just a one-liner per format, and the rest of the
workflow is identical. The chunk below is purely for illustration.

```{r, eval=FALSE}
jaspar.motifs   <- read_jaspar("JASPAR2024_CORE_vertebrates.txt")
hocomoco.motifs <- read_meme("HOCOMOCOv11_core_HUMAN_mono_meme.txt")
cisbp.motifs    <- read_cisbp("CIS-BP_2.00_Homo_sapiens.txt")
transfac.motifs <- read_transfac("transfac_matrix.dat")
uniprobe.motifs <- read_uniprobe("uniprobe_results.txt")
all.motifs      <- c(jaspar.motifs, hocomoco.motifs, cisbp.motifs,
                     transfac.motifs, uniprobe.motifs)
```

# Inspecting cross-source redundancy

Before deduplicating, it is worth quantifying how much redundancy
is actually in the combined set. `compare_motifs2()` in matrix
mode returns a square q-value matrix (one q-value per ordered
pair, after multiple-testing correction).
With several hundred motifs the result is a
`r format(length(all.motifs), big.mark = ",")`-row matrix that
holds every pairwise q-value. Two threads finish the comparison
in about four seconds on my laptop.

```{r}
qmat <- compare_motifs2(all.motifs,
                        matrix.out = "qvalue",
                        nthreads   = 2)
dim(qmat)
qvals <- qmat[upper.tri(qmat)]
length(qvals)
summary(qvals)
```

The pairwise q-value distribution is the visual cue for picking a
sensible deduplication threshold. On a log10 axis it spans many
orders of magnitude, with a long left tail of genuinely redundant
pairs (within-TF or cross-source duplicates) and a dense pile of
unrelated pairs heaped at q = 1.

```{r, fig.height=3, fig.width=6}
log10q <- log10(pmax(qvals, .Machine$double.xmin))
ggplot(data.frame(log10q = log10q), aes(x = .data$log10q)) +
  geom_histogram(binwidth = 0.5, fill = "black", colour = NA) +
  geom_vline(xintercept = log10(1e-6), linetype = "dashed",
             colour = "red") +
  labs(x = expression(log[10] * "(q-value) of pairwise comparison"),
       y = "Number of pairs") +
  theme_bw() +
  theme(panel.grid         = element_blank(),
        panel.border       = element_blank(),
        axis.line.x.bottom = element_line(colour = "black"),
        axis.line.y.left   = element_line(colour = "black"))
```

The dashed line marks q = 1e-6, the threshold we will use for the
deduplication in the next section. Picking that threshold is
something of a judgement call. Set it too loose and graph
transitivity starts collapsing unrelated motifs into giant
components (with this dataset, a threshold of q = 1e-3 produces a
single large cluster of hundreds of motifs, swallowing much of the
zinc-finger and homeodomain families, because their pairwise scores
share just enough small q-values to form one connected component).
Set it too tight and very few motifs are merged at all. For our
purposes here, q = 1e-6 seems to strike a reasonable balance,
collapsing a good deal of redundancy without bridging unrelated
families.

```{r}
mean(qvals < 1e-6)
sum(qvals < 1e-6)
```

# Deduplicating via graph clustering

`merge_similar2()` builds a graph whose edges link any two motifs
that pass the q-value threshold, finds the connected components,
and returns one representative motif per cluster. We use it twice:
once with `return.clusters = TRUE` to get the per-motif cluster
assignments (so we can confirm them visually), and once without, to
get the final merged set of motifs.

```{r}
clusters <- merge_similar2(all.motifs,
                           qvalue          = 1e-6,
                           return.clusters = TRUE,
                           nthreads        = 2)
```

To judge whether the q-value ought to be tightened further, we can
take a look at the largest cluster:

```{r, fig.height=6, fig.width=7}
big.cluster <- names(sort(table(clusters$cluster),
  decreasing = TRUE)[1])
big.cluster.motifs <- all.motifs[
  clusters$motif.i[clusters$cluster == big.cluster]]
if (length(big.cluster.motifs) > 10)
  big.cluster.motifs <- big.cluster.motifs[1:10]

view_motifs2(big.cluster.motifs,
  names.pos = "right", show.positions = FALSE)
```

It is well worth taking the time to go carefully back and forth
between examining the clusters and adjusting the q-value threshold,
since the final merge is really the most consequential decision in
curating a motif database. In our case we will carry on with this
threshold.

```{r}
merged <- merge_similar2(all.motifs,
                         qvalue   = 1e-6,
                         nthreads = 2)
length(all.motifs)
length(merged)
```

`merge_similar2()` autogenerates compound names for the merged
motifs by joining every contributing name with `+`, which again
becomes unwieldy once several motifs collapse into a single cluster.
A more compact alternative is to keep the first one or two
contributing names and roll the rest up into a count suffix, so that
the lineage stays visible without overflowing either the table or
the MEME header. Singletons simply keep their original name:

```{r}
contrib.names <- tapply(clusters$name, clusters$cluster,
                        function(n) as.character(n))
for (i in seq_along(merged)) {
  nms <- contrib.names[[i]]
  if (length(nms) == 2L)
    merged[[i]]["name"] <- paste(nms, collapse = "+")
  else if (length(nms) >= 3L)
    merged[[i]]["name"] <- sprintf("%s+%s+%dmore",
                                   nms[1], nms[2], length(nms) - 2L)
}
## `clusters` is a universalmotif_df: one row per input motif, with
## the motif objects in the `motif` column plus the `motif.i` and
## `cluster` bookkeeping columns. Show just the assignment columns:
print(head(as.data.frame(clusters)[, c("name", "motif.i", "cluster")], 8),
      row.names = FALSE)
```

The reduction is substantial, with roughly a third of the input
motifs collapsing into shared clusters. The distribution of cluster
sizes shows the long tail of singletons (motifs unique enough to
stay on their own), together with a moderate body of size-2 and
size-3 clusters (the common within-TF and cross-source
redundancies):

```{r, fig.height=3, fig.width=6}
cluster.sizes <- table(clusters$cluster)
size.dist     <- as.data.frame(table(cluster.sizes),
                               stringsAsFactors = FALSE)
names(size.dist) <- c("cluster.size", "n.clusters")
size.dist$cluster.size <- as.integer(size.dist$cluster.size)
ggplot(size.dist, aes(x = factor(.data$cluster.size),
                      y = .data$n.clusters)) +
  geom_col(fill = "black") +
  geom_text(aes(label = .data$n.clusters), vjust = -0.4, size = 3) +
  labs(x = "Cluster size (number of input motifs)",
       y = "Number of clusters") +
  theme_bw() +
  theme(panel.grid         = element_blank(),
        panel.border       = element_blank(),
        axis.line.x.bottom = element_line(colour = "black"),
        axis.line.y.left   = element_line(colour = "black"))
```

A cluster drawing on two or three different sources is direct
evidence of cross-database redundancy, which is precisely the kind
of duplication that a single-source library would quietly miss. We
can tally the sources per cluster:

```{r}
clusters$source <- vapply(all.motifs[clusters$motif.i],
                          function(m) m["family"], character(1))
sources.per.cluster <- tapply(clusters$source, clusters$cluster,
                              function(s) length(unique(s)))
table(sources.per.cluster)
```

Many clusters contain motifs from more than one source; those
multi-source clusters are exactly the cross-database redundancies
that this whole workflow exists to collapse.

# Visualising the cluster structure

`motif_tree2()` builds a hierarchical tree from the pairwise
comparison and renders it via `ggtree`. On all
`r format(length(all.motifs), big.mark = ",")` motifs the
labels would be too crowded to read, so we plot a random
200-motif subsample instead, coloured by source. The clades
typically mix all three sources, which mirrors the multi-source
cluster finding from above.

```{r, fig.height=6, fig.width=6, warning=FALSE, message=FALSE}
set.seed(2026)
sub.idx <- sort(sample(length(all.motifs), 200))
sub     <- all.motifs[sub.idx]
clean.names <- make.unique(gsub("[^A-Za-z0-9_]", "_",
                                vapply(sub, function(m) m["name"],
                                       character(1))))
for (i in seq_along(sub)) sub[[i]]["name"] <- clean.names[i]
motif_tree2(sub, linecol = "family", labels = "none",
            layout = "circular", tiplogo = FALSE, progress = FALSE) +
  ggplot2::scale_color_discrete(breaks = sources) +
  ggplot2::theme(legend.position = "bottom",
                 legend.title    = ggplot2::element_blank())
```

# Trimming uninformative flanks

PWM-style downstream scanning is sensitive to long low-information
flanks, which dilute scores and inflate match counts.
`trim_motifs()` removes columns whose information content falls
below `min.ic` (default 0.25 bits) from the edges. The default
"both" mode trims from both ends.

For a clean demonstration we pick the merged motif that loses the
most columns to trimming. Computing the trim delta directly (the
width before, minus the width after) is rather more reliable than
trying to guess from the flank IC, because `trim_motifs()` walks
inward from the edges and stops at the first column above `min.ic`,
which is not always the column the eye would have picked out as
"low IC".

```{r, fig.height=2.5, fig.width=7}
trim.delta <- vapply(merged, function(m)
  ncol(m["motif"]) - ncol(trim_motifs(m)["motif"]),
  integer(1))
trim.candidate.idx <- which.max(trim.delta)
trim.before <- merged[[trim.candidate.idx]]
trim.before["name"] <- paste0(trim.before["name"], "  (untrimmed)")
trim.after  <- trim_motifs(merged[[trim.candidate.idx]])
trim.after["name"] <- paste0(trim.after["name"], "  (trimmed)")
view_motifs2(list(trim.before, trim.after), names.pos = "right")
```

We then apply the trim to every merged motif before exporting:

```{r}
curated <- trim_motifs(merged)
range(sapply(merged,  function(m) ncol(m["motif"])))
range(sapply(curated, function(m) ncol(m["motif"])))
```

The width range shrinks, as expected: the merge step often pads
clusters out with low-confidence flanking columns wherever the input
motifs disagreed, and `trim_motifs()` is the natural way to clean
those back up.

# Adding the metadata back

For a curated library to be genuinely useful, users will want to
know, for each final motif, which sources contributed to it. The
`universalmotif_df` interface (`to_df()` returns a data-frame view,
and `update_motifs()` writes columns back into the motif list) is
the natural way to add per-motif annotations in a tidy fashion.

We compute the contributing-source list per cluster, then attach
it to the curated motifs together with a curation date:

```{r}
src.by.cluster <- tapply(clusters$source, clusters$cluster,
                         function(s)
                           paste(sort(unique(s)), collapse = ";"))
n.by.cluster   <- tapply(clusters$source, clusters$cluster,
                         length)

curated.df <- to_df(curated, extrainfo = TRUE)
curated.df$dataSource    <- NULL  # MotifDb's single-source field;
                                  # superseded by the multi-source
                                  # db.sources column we are about
                                  # to add
curated.df$db.sources    <- as.character(src.by.cluster)
curated.df$n.contributors <- as.integer(n.by.cluster)
curated.df$curation.date <- "2026-05-25"

print(as.data.frame(head(curated.df[, c("name", "db.sources",
                                        "n.contributors",
                                        "curation.date")], 10)),
      row.names = FALSE)
```

We round-trip back into motif objects with `update_motifs()`. The
new columns land in the `extrainfo` slot of each motif, and are then
preserved everywhere that `universalmotif` reads or writes its
native representation.

```{r}
curated <- to_list(update_motifs(curated.df), extrainfo = TRUE)
curated[[1]]["extrainfo"]
```

The MEME format has no slot for arbitrary key-value metadata, so how
much of the `extrainfo` survives is really limited by what the
output format itself can carry. For analyses where that metadata is
critical, my suggestion would be to write the curated library out as
a `universalmotif_df` RDS file (which preserves everything) and treat
the MEME export only as the format that consumers like FIMO and the
MEME suite happen to need.

# Export as a single MEME file

`write_meme()` writes the whole motif list out to a single file. The
MEME header carries the `name`, `altname`, `nsites`, and `eval` of
each motif, plus a global background.

```{r}
out.file <- tempfile(fileext = ".meme")
write_meme(curated, file = out.file, overwrite = TRUE)
file.info(out.file)$size
```

The first few lines of the file show the standard MEME preamble
and the start of the first motif block:

```{r}
cat(head(readLines(out.file), 25), sep = "\n")
```

We load it back with `read_meme()` to confirm that the file parses
cleanly and that the length matches:

```{r}
re.in <- read_meme(out.file)
length(re.in)
length(curated)
identical(sapply(re.in,   function(m) m["name"]),
          sapply(curated, function(m) m["name"]))
```

The names match, the motif matrices match, and the file is now
ready for any downstream tool that reads MEME-format motif
collections (FIMO, AME, CentriMo, or `read_meme()` itself for the
next analysis).

# Session info

```{r}
sessionInfo()
```

# References
