---
title: "Introduction to motif manipulation with the \"universalmotif\" package"
shorttitle: "Motif manipulation with \"universalmotif\""
author:
- name: Benjamin Tremblay
  affiliation: University of Waterloo, Waterloo, Canada
  email: b2tremblay@uwaterloo.ca
abstract: >
  Import, manipulate, and export motifs with R.
bibliography: universalmotif.bib
vignette: >
  %\VignetteIndexEntry{Introduciton to motif manipulation with the "universalmotif" package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::pdf_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Introduction to sequence motifs

# Importing and exporting motifs

## Importing

## Exporting

# Converting motif class

## Starting from another package

## Starting as a universalmotif motif

# Modifying motifs and related functions

## Accessing and changing universalmotif slots

## Converting motif type

## Comparing and merging motifs

## Multi-letter information

# Motif creation

## From a PCM/PPM/PWM/ICM matrix

## From sequences or character strings

## Generating random motifs

# Sequence-related functions

## Sequence shuffling and creation

## Searching for motifs in sequences

# Motif visualization

## Motif logos

## Motif trees with ggtree and motifStack

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

